Theories of deliberated judgment

### 2022-07-22 simple.tex
Clearer intro

\section{Introduction} 
Classical descriptive theories of choice in economics are limited to shallow observations.

This article defines theories that are prescriptive and empirical.
They are prescriptive in that the claim $a \succ b$ has the semantics of recommending choosing $a$ rather than $b$ when both are available.
They are empirical in that the validity of their recommendations rests on the basis of a mild axiom (compared to usual normative theories such as Savage’s Expected Utility theory) together with empirical claims, thus, claims that observations can falsify.

@@

\section{The model}
Let $\Phi = \set{\phi, ¬\phi}$ represent the decision at stake. Let $I$ be a set of individuals.

A prescriptive theory is a tuple $(\allargs, {\allleadsto})$.

$\allargs$ is a set of arguments. 
Throughout the article, $\N$ includes zero. %and $\N^* = \N \setminus \set{0}$. 
Define $\allhist = \bigcup_{k \in \N} \allargs^{\intvl{1, k}}$ as the finite sequences of arguments. A sequence of arguments is called an argumentative path, or simply a path.
A generic element of $\allhist$ is denoted by $\hist$ and called an argumentative path, or simply a path when this raises no ambiguity. Its length is denoted by $\card{\hist} \in \N$.
Given $j, l \in \N$, the notation $\intvl{j, l}$ represents $\set{k \in \N \suchthat j ≤ k ≤ l}$.
Thus, given an argumentative path $\hist \in \allhist$, $\forall k \in \intvl{1, \card{\hist}}$, $\hist_k \in A$, and $\hist = \emptyset ⇔ \card{\hist} = 0$.

@@

	Consider a set $\alts = \set{\alt_1, \alt_2} \subset \allalts$ of two items, and assume the two theories of \ac{DJ} are opposed on their prediction of the reflexive choice of individuals having to choose one item among $\alts$. Say, $\alt_1$ is satisfying and $\alt_2$ is not, failing on the speed criterion; while $v(\alt_2) > v(\alt_1)$.

>>>>

\section{Introduction} 
Observing subjects performing acts of choice is the basis of much work in econometrics. Such acts have been used, among others, to give an observable basis to the concept of preference, thanks in part to the seminal work of Samuelson. Acts of choice are considered here in the broad sense of choice between courses of actions, encompassing the simple setting of choosing between various elementary consumption bundles as well as settings involving more abstract acts such as choosing between defection or cooperation in a game-theoretical setting.

Most of these approaches consider only acts of choice that are performed spontaneously in the normal course of action of an individual, or that are given a specific frame depending on the psychological aspect that is to be studied. As an example of the former kind, typical willingness to pay studies asks the subject whether she prefers an object at some price or another object at another price, describing the objects in a way that is considered immediately understandable by the subject (or simply showing the objects). Examples of the latter kind include the celebrated work of Kahneman and Tversky.

On the other hand, since at least the seminal work of Fishkin, another sort of judgment involving subjective desirability is emerging as also worth studying. These are judgments given by subjects after arguments in favor and against different possible stance have been considered. Such judgments may differ from immediate judgments of desirability given in natural conditions. Indeed, individuals may change opinion on the desirability of some course of action, and therefore revise their choice, while learning about the properties of objects, the consequences of some acts (or empirical knowledge useful to estimate the likelihood of some consequences), or the logical or empirical impossibilities between consequences (for example, in some situations, you can’t both pay less taxes and have better government provided services). This is particularly likely to happen when choosing among non everyday objects or acts whose consequences are not well known by the subject.

While numerous articles have discussed the concept of deliberation in the last decades, it remains unsettled, to the best of my knowledge, how to observe a deliberated judgment, or, correspondingly, how to define this concept in a way that makes it amenable to empirical study, sufficiently precisely that any disagreement about someone’s deliberated judgment in a given context could be solved, at least in principle, by an empirical experiment. %This article is interested in defining the concept of deliberated judgementof a subject, as opposed to the classical one, that will henceforth be called her natural judgment, that is, the way that she chooses between acts when she does not face any arguments, i.e.,

While the concept of deliberation in the deliberative democracy literature usually refers to multiple individuals debating, this article considers deliberation as related to a single individual at a time: an individual deliberates (in the sense of: carefully weights) the stance she considers most appropriate while being confronted to arguments. The deliberated judgment of an individual is thus defined with no requirement that the arguments be uttered by peers in a debate; they can be given by an automata, for example. In this sense, deliberated judgments come close to what Rawls calls reflective equilibrium (referring to an idea of Goodman), a concept that has much inspired the present proposal.

To define deliberated judgments, one needs to start with the set of arguments under consideration, and the way the subject gets to know them, called hereafter the “exposure protocol”. As a result, we need always talk about the deliberated judgment \emph{given some exposure protocol}. The notion of argument invoked here is an extremely large one, including texts, images, sounds, experiences (a hiking trip), and so on: anything that can be transmitted to a subject qualifies as a possible argument. Such an “argument” need not even have the logical structure of what is considered a proper argument in argumentation theory, which permits to avoid having to take position on what is a correct argument, and permits to study wide enough conceptions of deliberated judgment, including those involving no paternalism: anything that influences the judgment of a subject may a priori be considered worth including in at least some conceptions of deliberated judgment.

The dependence of the deliberated judgment concept on the exposure protocol yields a first qualification about the concept of deliberated judgments defined here when one desires to use it normatively: it is only as reasonable to call the resulting judgments “deliberate” in a normative sense as the exposure protocol is normatively relevant to deliberation. A sufficiently restricted, or even purposefully chosen, set of arguments may yield judgments that are only deliberate in the technical sense used here, but that pertain more to brain-washing than to reflective judgments. While it is indeed a long-term goal of this research stream to yield a normatively relevant conception of preference (appropriate, for example, for recommendation), it is not a claim of this article that anything that corresponds to deliberated judgments as defined here qualify as being appropriate for recommendation (or for any other normative use): much depends on the chosen exposure protocol. The goal of this article is, more modestly, one of a conceptual and practical clarification. It proposes a sharp distinction between the empirical content and the normative content of the concept of deliberated judgments: the definition proposed here transfers the full normative weight on the choice of the exposure protocol; and leaves the rest of the disagreements to empirical settling. This is methodologically akin, mutatis mutandis, to the axiomatic study of justice such as done in social choice, only replacing mathematical deduction with empirical claims. In social choice, the choice of axioms bears the full normative weight; once axioms are chosen, it is a matter of deduction which social choice rules (if any) correspond to that choice. In our case, once an exposure protocol is chosen, it is a pure empirical matter which deliberated judgments result.
In supplement to a possibly helpful conceptual clarification, such a sharp separation may reveal practically useful for separation of concern as it permits to study deliberated judgments empirically under various exposure protocols independently of possible persistent disagreements about which norms are more appropriate for deliberation in which circumstances.

The deliberated judgment of a subject is said to depend on the exposure protocol, that is, on the set of arguments and the way they are submitted to the subject, and not only on the set of arguments, because the very way that they are transmitted to the individual may affect her judgment differently. For example, reading an image description and showing an image to a subject may impress them differently.

The exposure protocol defines a set of argument and, importantly, how arguments are sent to the subject and her subsequent choice observed. There is, as need be, no requirement on the internal process that occurs in the subject’s mind while she (possibly) processes the arguments, as it is not desirable to assume that the internal process could be observed. We consider subjects as black boxes and only require a capacity of observation of their choices after exposure to arguments. Depending on the exposure protocol and the subject, it is of course possible for the arguments to never affect any judgment of the subject, a conclusion that may be interesting in its own right.

The deliberated judgment of a subject is defined as a function of the exposure protocol, as accepting every claims that resists every counter-arguments. The notion of resistance is itself defined according to the exposure protocol.

Studying judgments resulting from exposure to arguments includes a difficulty that is not present with natural choice experiments: exposing an individual to arguments may change the individual’s stance, an effect that cannot be erased to submit the individual (considered “clean” again) to another sequence of arguments. The observer thus cannot in general assume that multiple unrelated sequences of arguments can be tested on a given individual. This difficulty will be worked around by considering general theories of deliberated judgments, that apply to sets of individuals. This is methodologically similar to defining theories of how objects break. One can test only one way of breaking a given object (assuming it cannot be repaired), so such theories must apply to sets of objects if one wants to be able to test multiple ways of falsifying the theories.

@@

\section{The model}
Let $\Phi = \set{\phi, ¬\phi}$ represent the decision at stake.

An exposure protocol is a tuple $(I, \allargs, {\allleadsto})$, where $I$ is a set of individuals, $\allargs$ is a set of arguments, and ${\allleadsto}$ is the behavior of the individuals as a function of the arguments they have been exposed to.

$\allargs$ is a set of arguments. 
Throughout the article, $\N$ includes zero. %and $\N^* = \N \setminus \set{0}$. 
Given $j, l \in \N$, the notation $\intvl{j, l}$ represents $\set{k \in \N \suchthat j ≤ k ≤ l}$ (thus $\forall j \in \N: \intvl{j, 0} = \emptyset$).
Define $\allhist = \bigcup_{k \in \N} \allargs^{\intvl{1, k}}$ as the finite sequences of arguments. A sequence of arguments is called an argumentative path, or simply a path.
A generic element of $\allhist$ is denoted by $\hist$ and called an argumentative path, or simply a path when this raises no ambiguity. Its length is denoted by $\card{\hist} \in \N$.
Thus, given an argumentative path $\hist \in \allhist$, $\forall k \in \intvl{1, \card{\hist}}$, $\hist_k \in \allargs$, and $\hist = \emptyset ⇔ \card{\hist} = 0$.

@@

	Consider a set $\alts = \set{\alt_1, \alt_2} \subset \allalts$ of two items, and assume the two theories of \ac{DJ} are opposed on their prediction of the reflective choice of individuals having to choose one item among $\alts$. Say, $\alt_1$ is satisfying and $\alt_2$ is not, failing on the speed criterion; while $v(\alt_2) > v(\alt_1)$.
